---
title: "F Test Power Simulation"
author: "Jingyang(Judy) Zhang"
date: "10/25/2021"
output: html_document
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```



```{r load_pkg}
library(tidyverse)
library(ggpubr)
```

# Intro

In this simulation, we will try different combinations of values of the three parameters:

$n_0$: control group sample size; 

$n_1$: treatment group sample size;

$\phi = \frac{\sigma_1^2}{\sigma_0^2}$: ratio of population variances of the treatment group over the control group.


The power can be calculated by:

$Pr(F > \phi F_{n_1 -1, n_0 -1})$ 

where $F = \frac{s_1^2}{s_0^2} = \frac{MSE_1}{MSE_0}$



# Functions

Function: cal_power() calculates power for a two-sided F-test for equality of two variances. 

Arguments: 
var1_popl: population variance for the treatment group.
var0_popl: population variance for the control group. 
var1: sample variance for the treatment group.
var0 sample variance for the control group. 
n1: sample size for the treatment group.
n0: sample size for the control group. 
alpha: significance level for the F-test of equality of two population variances. 

Return:
power: calculated power for the F-test for equality of two variances. 


```{r cal_power}

cal_power <- function(var1_popl, var0_popl, var1, var0, n1, n0, alpha){
  

  
  F_stat <- var1/var0
  
  
  phi <- var1_popl/var0_popl
  
  F_crit_low <- qf(alpha/2, n1-1, n0-1, lower.tail = TRUE, log.p = FALSE)
  
  F_crit_up <- qf(1 - alpha/2, n1-1, n0-1, lower.tail = TRUE, log.p = FALSE)
  
  p_low <- pf(phi*F_crit_low, n1-1, n0-1, lower.tail = TRUE, log.p = FALSE)
  
  p_up <- pf(phi*F_crit_up, n1-1, n0-1, lower.tail = FALSE, log.p = FALSE)
  
  power <- p_low + p_up
  
  return(power)
  
  
}


```


Test cal_power() Function

Assume $X\sim N(\mu_0, \sigma_0^2)$ is the control group population and $Y\sim N(\mu_1, \sigma_1^2)$ is the treatment group population.

We will be randomly drawing a sample of size n0 from the control group population and a sample of size n1 from the treatment group population. 



Test Data:
$X\sim N(0, 1)$ 

$Y\sim N(0, 10)$  

$n_0 = n_1 = 30$

```{r test_cal_power}

var1_popl <- 180
var0_popl <- 175
mu1 <- 0
mu0 <- 0 


n1 <- 30
n0 <- 30


trt_sample <- rnorm(n1, mean = mu1, sd = sqrt(var1_popl))
ctrl_sample <- rnorm(n0, mean = mu0, sd = sqrt(var0_popl))

var1 <- (sd(trt_sample))^2
var0 <- (sd(ctrl_sample))^2

cal_power(var1_popl, var0_popl, var1, var0, n1, n0, 0.05)

```



Function: simulate_power()

Arguments:

var1popl_list: a vector containing a list of population variances for different treatment groups. 

var0popl_list: a vector containing a list of population variances for different control groups. 

var1_list: a vector containing a list of sample variances for different treatment group samples. 

var0_list: a vector containing a list of sample variances for different control group samples. 

n1_list: a vector containing a list of sample sizes for different treatment group samples. 

n0_list: a vector containing a list of sample sizes for different treatment group samples. 

alpha_list: a vector containing a list of significance levels. 


Return: 
result_table: a tibble containing parameters and corresponding power of the F test. 



```{r simulate_power}

simulate_power <- function(var1popl_list, var0popl_list, var1_list, var0_list, n1_list, n0_list, alpha_list){
  
  power <- c()
  
  for(i in 1:length(var1popl_list)){
    
    power[i] <- cal_power(var1popl_list[i], var0popl_list[i], var1_list[i], var0_list[i], n1_list[i], n0_list[i], alpha_list[i])
    
  }
  
  # Result table will provide three parameters and corresponding power: phi, n1, n0, and power
  
  result_table <- tibble(phi = var1popl_list/var0popl_list, n1 = n1_list, n0 = n0_list, power = power)
  
  return(result_table)
  
  
}


```



Test Data:


```{r test_simulate_power}

var1popl_list <- c(10, 20)
var0popl_list <- c(15, 21)
mu1 <- 0
mu0 <- 0 


n1_list <- c(30, 30)
n0_list <- c(30, 30)


var1_list <- c()
var0_list <- c()

for(i in 1:length(var1popl_list)){
  
  trt <- rnorm(n1_list[i], mean = mu1, sd = sqrt(var1popl_list[i]))
  var1_list[i] <- (sd(trt))^2
  ctrl <- rnorm(n0_list[i], mean = mu0, sd = sqrt(var0popl_list[i]))
  var0_list[i] <- (sd(ctrl))^2
  
}

alpha_list <- c(0.05, 0.05)


result_table <- simulate_power(var1popl_list, var0popl_list, var1_list, var0_list, n1_list, n0_list, alpha_list)


```



# Simulations

## Same Sample Size with Different Phi

```{r simulate_phi}


var1popl_list <- seq(10, 1, by = - 0.01)
var0popl_list <- rep(10, length(var1popl_list))
mu1 <- 0
mu0 <- 0 


n1_list <- rep(30, length(var1popl_list))
n0_list <- rep(30, length(var1popl_list))


var1_list <- c()
var0_list <- c()

for(i in 1:length(var1popl_list)){
  
  trt <- rnorm(n1_list[i], mean = mu1, sd = sqrt(var1popl_list[i]))
  
  var1_list[i] <- (sd(trt))^2
  
  ctrl <- rnorm(n0_list[i], mean = mu0, sd = sqrt(var0popl_list[i]))
  
  var0_list[i] <- (sd(ctrl))^2
  
}

alpha_list <- rep(0.05, length(var1popl_list))


result_table_phi <- simulate_power(var1popl_list, var0popl_list, var1_list, var0_list, n1_list, n0_list, alpha_list)


```





Plot phi v.s. power

```{r plot_phi}

ggplot(result_table_phi, aes(x = phi, y = power)) + 
  geom_point() + 
  scale_x_continuous(breaks = seq(0, 1, by = 0.05)) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.05))
  theme(axis.text.x = element_text(angle = 295, vjust = 0.5, hjust=1))

```

1. As $\phi\rightarrow 1$, power declines.

2. As $\phi\rightarrow 1$, the rate of decline of the power increases and then decreases.

3. As $\phi\rightarrow 1$, the power decreases faster for $\phi\in [0.30, 0.60]$ than for $\phi\in (0, 0.30)\cup(0.6, 1)$.




## Different Sample Sizes with the Same Phi

```{r simulate_sample_sizes}

n1_list <- seq(10, 100, by = 1)
n0_list <- rep(100, length(n1_list))

var1popl_list <- rep(5, length(n1_list))
var0popl_list <- rep(10, length(n1_list))
mu1 <- 0
mu0 <- 0 





var1_list <- c()
var0_list <- c()

for(i in 1:length(var1popl_list)){
  
  trt <- rnorm(n1_list[i], mean = mu1, sd = sqrt(var1popl_list[i]))
  
  var1_list[i] <- (sd(trt))^2
  
  ctrl <- rnorm(n0_list[i], mean = mu0, sd = sqrt(var0popl_list[i]))
  
  var0_list[i] <- (sd(ctrl))^2
  
}

alpha_list <- rep(0.05, length(n1_list))


result_table_size <- simulate_power(var1popl_list, var0popl_list, var1_list, var0_list, n1_list, n0_list, alpha_list)


```



Plot sample size v.s. power

```{r plot_sample_size}

ggplot(result_table_size, aes(x = n1/n0, y = power)) + 
  geom_point() + 
  scale_x_continuous(breaks = seq(0, 1, by = 0.05)) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.05))
  theme(axis.text.x = element_text(angle = 295, vjust = 0.5, hjust=1))

```

1. As $\frac{n_1}{n_0}\rightarrow 1$, power increases. 

2. As $\frac{n_1}{n_0}\rightarrow 1$, the rate of increase of power decreases. 





## Sample Size, Phi and Power


### $\phi = 0.95$
```{r simulate_samplesize_phi95}


n1_list <- seq(10, 200, by = 1)
n0_list <- n1_list

var1popl_list <- rep(9.5, length(n1_list))
var0popl_list <- rep(10, length(n1_list))
mu1 <- 0
mu0 <- 0 





var1_list <- c()
var0_list <- c()

for(i in 1:length(var1popl_list)){
  
  trt <- rnorm(n1_list[i], mean = mu1, sd = sqrt(var1popl_list[i]))
  
  var1_list[i] <- (sd(trt))^2
  
  ctrl <- rnorm(n0_list[i], mean = mu0, sd = sqrt(var0popl_list[i]))
  
  var0_list[i] <- (sd(ctrl))^2
  
}

alpha_list <- rep(0.05, length(n1_list))


result_table_phi95 <- simulate_power(var1popl_list, var0popl_list, var1_list, var0_list, n1_list, n0_list, alpha_list)



```




Plot sample size v.s. power for $\phi = 0.95$

```{r plot_samplesize_phi95}

phi95 <- ggplot(result_table_phi95, aes(x = n1, y = power)) + 
  geom_point() + 
  scale_x_continuous(breaks = seq(5, 210, by = 10)) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.001)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  
phi95

```

1. As $n_1 = n_0 = n\rightarrow 1$, power increases. 

2. As $n_1 = n_0 = n\rightarrow 1$, the rate of increase of power decreases. 





### $\phi = 0.9$
```{r simulate_samplesize_phi90}


n1_list <- seq(10, 200, by = 1)
n0_list <- n1_list

var1popl_list <- rep(9, length(n1_list))
var0popl_list <- rep(10, length(n1_list))
mu1 <- 0
mu0 <- 0 





var1_list <- c()
var0_list <- c()

for(i in 1:length(var1popl_list)){
  
  trt <- rnorm(n1_list[i], mean = mu1, sd = sqrt(var1popl_list[i]))
  
  var1_list[i] <- (sd(trt))^2
  
  ctrl <- rnorm(n0_list[i], mean = mu0, sd = sqrt(var0popl_list[i]))
  
  var0_list[i] <- (sd(ctrl))^2
  
}

alpha_list <- rep(0.05, length(n1_list))


result_table_phi90 <- simulate_power(var1popl_list, var0popl_list, var1_list, var0_list, n1_list, n0_list, alpha_list)



```




Plot sample size v.s. power for $\phi = 0.9$

```{r plot_samplesize_phi90}

phi90 <- ggplot(result_table_phi90, aes(x = n1, y = power)) + 
  geom_point() + 
  scale_x_continuous(breaks = seq(5,230, by = 10)) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.01)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  
phi90

```

1. As $n_1 = n_0 = n\rightarrow 1$, power increases. 

2. As $n_1 = n_0 = n\rightarrow 1$, the rate of increase of power decreases. 




### $\phi = 0.7$
```{r simulate_samplesize_phi70}


n1_list <- seq(10, 200, by = 1)
n0_list <- n1_list

var1popl_list <- rep(7, length(n1_list))
var0popl_list <- rep(10, length(n1_list))
mu1 <- 0
mu0 <- 0 





var1_list <- c()
var0_list <- c()

for(i in 1:length(var1popl_list)){
  
  trt <- rnorm(n1_list[i], mean = mu1, sd = sqrt(var1popl_list[i]))
  
  var1_list[i] <- (sd(trt))^2
  
  ctrl <- rnorm(n0_list[i], mean = mu0, sd = sqrt(var0popl_list[i]))
  
  var0_list[i] <- (sd(ctrl))^2
  
}

alpha_list <- rep(0.05, length(n1_list))


result_table_phi70 <- simulate_power(var1popl_list, var0popl_list, var1_list, var0_list, n1_list, n0_list, alpha_list)



```




Plot sample size v.s. power for $\phi = 0.7$

```{r plot_samplesize_phi70}

phi70 <- ggplot(result_table_phi70, aes(x = n1, y = power)) + 
  geom_point() + 
  scale_x_continuous(breaks = seq(5, 210, by = 10)) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.05)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  
phi70

```

1. As $n_1 = n_0 = n\rightarrow 1$, power increases. 

2. As $n_1 = n_0 = n\rightarrow 1$, the rate of increase of power decreases. 



### $\phi = 0.5$
```{r simulate_samplesize_phi50}


n1_list <- seq(10, 200, by = 1)
n0_list <- n1_list

var1popl_list <- rep(5, length(n1_list))
var0popl_list <- rep(10, length(n1_list))
mu1 <- 0
mu0 <- 0 





var1_list <- c()
var0_list <- c()

for(i in 1:length(var1popl_list)){
  
  trt <- rnorm(n1_list[i], mean = mu1, sd = sqrt(var1popl_list[i]))
  
  var1_list[i] <- (sd(trt))^2
  
  ctrl <- rnorm(n0_list[i], mean = mu0, sd = sqrt(var0popl_list[i]))
  
  var0_list[i] <- (sd(ctrl))^2
  
}

alpha_list <- rep(0.05, length(n1_list))


result_table_phi50 <- simulate_power(var1popl_list, var0popl_list, var1_list, var0_list, n1_list, n0_list, alpha_list)



```




Plot sample size v.s. power for $\phi = 0.5$

```{r plot_samplesize_phi50}

phi50 <- ggplot(result_table_phi50, aes(x = n1, y = power)) + 
  geom_point() + 
  scale_x_continuous(breaks = seq(5, 210, by = 10)) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.05)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  
phi50

```

1. As $n_1 = n_0 = n\rightarrow 1$, power increases. 

2. As $n_1 = n_0 = n\rightarrow 1$, the rate of increase of power decreases. 




Compare

```{r plots_samplesize_phi}

ggarrange(phi95, phi90, phi70, phi50,
          labels = c("phi = 0.95", "phi = 0.9", "phi = 0.7", "phi = 0.5"),
          label.x = 0.5,
          label.y = 0.5,
          font.label = list(size = 12),
          ncol = 2, nrow = 2)




```


1. One obvious pattern is that for any given sample size n, the larger the $\phi$, the smaller the power. 


2. For large value of $\phi$, the effect of increasing sample size n on increasing the power is about the same for small and large sample sizes. 

3. For small value of $\phi$, there is a diminishing effect of increasing sample size n on increasing the power. That is, as sample size gets larger, the increase in power gets smaller. 


In general, as $\phi$ decreases, the relationship between sample size n and power goes from a linear function to a concave function. 

