---
title: "Relate_Two_F_Tests"
author: "Jingyang(Judy) Zhang"
date: "11/11/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Load Packages and External R Scripts
```{r load_pkg_source}
source("F_Test_Power_Mean_Functions.R")
source("F_Test_Power_Variance_Functions.R")
library(pwr)
library(tidyverse)
```



# Introduction

## Setup

Treatment Group Sample Observations: $Y_{1j}\stackrel{iid}\sim N(\mu_{1}, \sigma_{1}^{2})$, $j = 1, ..., n_{1}$.

Control Group Sample Observations: $Y_{0j}\stackrel{iid}\sim N(\mu_{0}, \sigma_{0}^{2})$, $j = 1, ..., n_{0}$.

Total Number of Groups: $k = 2$

Total Number of Observations: $N = \sum_{i=1}^{k}n_{i} = 2n$


Effect Size:

We will use standardized mean difference ($\delta$) as the effect size:

$\delta = \frac{\mu_{1} - \mu_{0}}{\sigma}$

**Assumptions:**

1. Two populations (i.e. treatment group and control group) are independent.

2. Assume equal sample size: $n_{0} = n_{1} = n$.



# Calculate Sample Sizes

**We are going to calculate sample sizes required for a F-Test for equality of population means (which is equivalent to a two sample t-test for equality of population means when there are only two independent groups) to obtain a power of 80%, given $\alpha = 0.5$ and a set of different effect sizes (standardized mean differences: $\delta = \frac{\mu_{1} - \mu_{0}}{\sigma}).**

**Assumption:**

1. Assume equal population variances $\sigma^{2}_{0} = \sigma^{2}_{1} = \sigma^{2}$.

2. Assume we are conducting a two-sided F-test/t-Test for equality of population means. 

$H_{0}: \mu_{0} = \mu_{1}$

$H_{a}: \mu_{0}\neq \mu_{1}$


**Test Data:**


$\sigma^{2} = 1$

$\mu_{0} = 0, 0, 0, 0, 0$

$\mu_{1} = 0.1, 0.2, 0.3, 0.4, 0.5$

The above setup will result in effect sizes $\delta = \frac{\mu_{1} - \mu_{0}}{\sigma} = 0.1, 0.2, 0.3, 0.4, 0.5$

$\alpha = 0.05$

$\text{Power} = 0.8$

```{r cal_sample_size_mean}

ds <- seq(0.1, 0.5, by = 0.1)

ns_mean <- c()

for(i in 1:length(ds)){
  
  n_mean <- round(as.numeric(pwr.t.test(d=ds[i], sig.level=0.05, power=0.80, type="two.sample", alternative="two.sided")$n))
  
  ns_mean[i] <- n_mean

}




```






# Calculate Power

**We are going to calculate power of a F-Test for equality of population variances, given $\alpha = 0.05$, a set of different population variances ratios $\phi = \frac{\sigma^{2}_{1}}{\sigma^{2}_{0}}$, and a set of different effect sizes (standardized mean differences: $\delta = \frac{\mu_{1} - \mu_{0}}{\sigma}).**



**Assumption:**

1. Note: we no longer assume equal population means. 

2. Assume we are conducting a two-sided F-test/t-Test for equality of population variances

$H_{0}: \frac{\sigma^{2}_{1}}{\sigma^{2}_{0}} = 1\Rightarrow \sigma^{2}_{0} = \sigma^{2}_{1}$

$H_{a}: \frac{\sigma^{2}_{1}}{\sigma^{2}_{0}} \neq 1\Rightarrow \sigma^{2}_{0} \neq \sigma^{2}_{1}$


**Test Data:**


$\sigma^{2}_{0} = 1, 1, 1, 1, 1$

$\sigma^{2}_{1} = 1.1, 1.2, 1.3, 1.4, 1.5, $

$\mu_{0} = 0, 0, 0, 0, 0$

$\mu_{1} = 0.1, 0.2, 0.3, 0.4, 0.5$

The above setup will result in effect sizes $\delta = \frac{\mu_{1} - \mu_{0}}{\sigma} = 0.1, 0.2, 0.3, 0.4, 0.5$

The above setup will result in population variance ratios $\phi = \frac{\sigma^{2}_{1}}{\sigma^{2}_{0}}} = 1.1, 1.2, 1.3, 1.4, 1.5$

$\alpha = 0.05$

We will look at power of the F-Test for equality of population variances using sample sizes (per group) calculated above:

$n_{0} = n_{1} = n = 1571, 393, 175, 99, 64$



## Power for $\phi = 1.1$, $\delta = 0.1$

```{r cal_power_variance}


sigma_sqr1s <- rep(1.1, length(ns_mean))
sigma_sqr0s <- rep(1, length(ns_mean))

mu1 <- 0.1
mu0 <- 0 

alphas <- rep(0.05, length(ns_mean))



s_sqr1s <- c()
s_sqr0s <- c()





for(i in 1:length(ns_mean)){
  
  trt <- rnorm(ns_mean[i], mean = mu1, sd = sqrt(sigma_sqr1s[i]))
  
  s_sqr1s[i] <- (sd(trt))^2
  
  ctrl <- rnorm(ns_mean[i], mean = mu0, sd = sqrt(sigma_sqr0s[i]))
  
  s_sqr0s[i] <- (sd(ctrl))^2
  
}




result_table <- simulate_power_var(sigma_sqr1s, sigma_sqr0s, s_sqr1s, s_sqr0s, ns_mean, ns_mean, alphas)

result_table %>% add_column(delta = rep(0.1, 5), .before = "phi")











```
